Follow-up Samples

### 2017-05-13 submission/manuscript.tex
fix R code comments

\textbf{Conclusions.} Further exploration of the relationship between
diagnosis, treatment, and the impact on the microbiota will yield
improvements in disease management.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiota of patients with adenoma (N=22),
advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
between pre and post-treatment samples {[}Table 1{]}. The structure of
the microbial communities of the pre and post-treatment samples
differed, as measured by the \(\theta\)\textsubscript{YC} beta diversity
metric {[}Figure 1A{]}. We found that the communities obtained pre and
post-treament among the patients with carcinomas changed significantly
more than those patients with adenoma (P-value = 5.4e-05). There were no
significant differences in the amount of change observed between the
patients with adenoma and advanced adenoma or between the patients with
advanced adenoma and carcinoma (P-value \textgreater{} 0.05). Next, we
tested whether there was a consistent direction in the change in the
community structure between the pre and post-treatment samples for each
of the diagnosis groups {[}Figure 1B-D{]}. We only observed a consistent
shift in community structure for the patients with carcinoma when using
a PERMANOVA test (adenoma P-value =0.999, advanced adenoma P-value
=0.945, and carcinoma P-value =0.005). Finally, we measured the number
of observed OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value NA
0.05). It is likely that high inter-individual variation and the
irregular distribution of OTUs across individuals limited the
statistical power of the test. To overcome these problems we developed
Random Forest models to identify collections of OTUs that would allow us
to differentiate between pre and post-treatment samples from each of the
diagnosis groups. Because of the relatively small number of subjects in
each group, we restricted our models to only incorporate 10 OTUs to
limit the likelihood that the models would overfit the data. Despite
this restriction, the models performed well (adenoma AUC range = 0.69 -
0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
had little overlap with each other {[}Figure 2{]}. These results support
the earlier community-wide analysis where we observed that the treatment
had an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

@@

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. In addition, it is possible that the difference
between the microbiota of patients with adenoma and advanced adenoma and
those with normal colons is subtle. This is supported by the reduced
ability of our models to correctly classify patients with adenomas and
advanced adenomas relative to those diagnosed with carcinomas {[}Figure
S1{]}. Given the irregular distribution of microbiota across patients in
the different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

There was a subset of patients (6 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 23.08\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 6 individuals (3 with Stage II and 3 with Stage III) was
similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased (382 \(\pm\) 116 days) or decreased (341 \(\pm\)
98 days) (Wilcoxon Test; P-value = 0.56). Finally, it is possible that
these patients may not have responded to treatment as well as the other
20 patients diagnosed with carcinoma and so the microbiota may not have
been impacted the same way. Again, further studies looking at the role
of the microbiota in recurrence are needed to understand the dynamics
following treatment.

Our final hypothesis was that the specific type of treatment altered the
structure of the microbiome. The treatment to remove adenomas and
advanced adenomas was either polyp removal or surgical resection whereas
it was surgical resection alone or in combination with chemotherapy or
with chemotherapy and radiation for individuals with carcinoma. Because
chemotherapy and radiation target rapidly growing cells, these
treatments would be more likely to cause a turnover of the colonic
epithelium driving a more significant change in the structure of the
microbiota. Although, we were able to test for an effect across these
specific types of treatment, the number of patients in each treatment
group was relatively small. Finally, those undergoing surgery would have
received antibiotics and this may be a potential confounder. However,
our pre-treatment stool samples were obtained before the surgery and the
post-treatment samples were obtained long after any effects due to
antibiotic administration on the microbiome would be expected to occur
(351 \(\pm\) 102 days).

@@

\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
used to create the three models used. The adenoma model classified
normal versus adenoma, the advanced adenoma model classified normal
versus advanced adenoma, and the carcinoma model classified normal
versus carcinoma. The total number of individuals in the treatment
models was 67 individuals. There were a total of 22 individuals in the
treatment adenoma model, 19 individuals in the treatment advanced
adenoma model, and 26 individuals in the treatment carcinoma model
{[}Table 1{]}. The total number of individuals in the normal versus
diagnosis models was 423 individuals {[}Table 2{]}. There were a total
of 239 individuals in the adenoma model, 262 individuals in the advanced
adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
Each model was then applied to our 67-person cohort {[}Table 1{]} which
assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
(advanced adenoma n = 0 and disease free n = 19), pre-treatment
carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
carcinoma (carcinoma n = 1 and disease free n = 25).

The model included only OTU data obtained from 16S rRNA sequencing.
Non-binary data was checked for near zero variance and OTUs that had
near zero variance were removed. This pre-processing was performed with
the R package caret (v6.0.73). Optimization of the mtry hyper-parameter
involved making 100 different 80/20 (train/test) splits of the data
where normal and adenoma, normal and advanced adenoma, or normal and
carcinoma were represented in the same proportion within both the whole
data set and the 80/20 split. For each of the different splits, 20
repeated 10-fold cross validation was performed on the 80\% component to
optimize the mtry hyper-parameter by maximizing the AUC (Area Under the
Curve of the Receiver Operator Characteristic). The resulting model was
then tested on the hold out data obtained from the 20\% component. All
three models had an optimized mtry of 2. The hyper-parameter, mtry,
defines the number of variables to investigate at each split before a
new division of the data was created with the Random Forest model.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre and post-treatment samples utilized a paired
Wilcoxon ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}35{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtaining the median
MDA from the 100, 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
and Carcinoma Groups After Treatment.} A) Thetayc distance from pre
versus post sample within each individual. A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
5.36e-05). Solid black points represent the median value for each
diagnosis group. B) NMDS of the pre and post-treatment samples for the
adenoma group. C) NMDS of the pre and post-treatment samples for the
advanced adenoma group. D) NMDS of the pre and post-treatment samples
for the carcinoma group.

\textbf{Figure 2: The 10 OTUs used to classify treatment for Adenoma,
Advanced Adenoma, and Carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs.

\textbf{Figure 3: Common OTUs to All Models.} A) Venn diagram showing
the OTU overlap between each model. B) For each common OTU the lowest
taxonomic identification and importance rank for each model run is
shown.

\textbf{Figure 4: Treatment Response Based on Models Built for Adenoma,
Advanced Adenoma, or Carcinoma.} A) Positive probability change from
initial to follow up sample in those with adenoma. B) Positive
probability change from initial to follow up sample in those with
advanced adenoma. C) Positive probability change from initial to follow
up sample in those with carcinoma.

@@

\textbf{Table 1: Demographic Data of pre and Post Treatment Cohort}

@@

\textbf{Table 2: Demographic Data of Training Cohort}

@@

\textbf{Figure S1: ROC Curves of the Adenoma, Advanced Adenoma, and
Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
represent the range of values of a 100 different 80/20 splits of the
test set data and the dark green line represents the model using 100\%
of the data set and what was used for subsequent classification. B)
Advanced Adenoma ROC curve: The light yellow shaded areas represent the
range of values of a 100 different 80/20 splits of the test set data and
the dark yellow line represents the model using 100\% of the data set
and what was used for subsequent classification. C) Carcinoma ROC curve:
The light red shaded areas represent the range of values of a 100
different 80/20 splits of the test set data and the dark red line
represents the model using 100\% of the data set and what was used for
subsequent classification.

\textbf{Figure S2: Summary of Important OTUs for the Adenoma, Advanced
Adenoma, and Carcinoma Models.} A) MDA of the most important variables
in the adenoma model. The dark green point represents the mean and the
lighter green points are the value of each of the 100 different runs. B)
Summary of Important Variables in the advanced adenoma model. MDA of the
most important variables in the SRN model. The dark yellow point
represents the mean and the lighter yellow points are the value of each
of the 100 different runs. C) MDA of the most important variables in the
carcinoma model. The dark red point represents the mean and the lighter
red points are the value of each of the 100 different runs.

\textbf{Figure S3: Pre and Post Treatment Relative Abundance of CRC
associated OTUs within the Carcinoma Model.}

>>>>

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will be validated that can be used to quantify the risk of
recurrence and the likelihood of survival.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiota of patients with adenoma (N = 22),
advanced adenoma (N = 19), or carcinoma (N = 26) had any broad
differences between pre and post-treatment samples {[}Table 1{]}. The
structure of the microbial communities of the pre and post-treatment
samples differed, as measured by the \(\theta\)\textsubscript{YC} beta
diversity metric {[}Figure 1A{]}. We found that the communities obtained
pre and post-treament among the patients with carcinomas changed
significantly more than those patients with adenoma (P-value \textless{}
0.001). There were no significant differences in the amount of change
observed between the patients with adenoma and advanced adenoma or
between the patients with advanced adenoma and carcinoma (P-value
\textgreater{} 0.05). Next, we tested whether there was a consistent
direction in the change in the community structure between the pre and
post-treatment samples for each of the diagnosis groups {[}Figure
1B-D{]}. We only observed a consistent shift in community structure for
the patients with carcinoma when using a PERMANOVA test (adenoma P-value
= 0.999, advanced adenoma P-value = 0.945, and carcinoma P-value =
0.005). Finally, we measured the number of observed OTUs, Shannon
evenness, and Shannon diversity in the pre and post-treatment samples
and did not observe a significant change for any of the diagnosis groups
(P-value \textgreater{} 0.05) {[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups (P-value
\textgreater{} 0.05). It is likely that high inter-individual variation
and the irregular distribution of OTUs across individuals limited the
statistical power of the test. To overcome these problems we developed
Random Forest models to identify collections of OTUs that would allow us
to differentiate between pre and post-treatment samples from each of the
diagnosis groups. To limit the likelihood that the models would overfit
the data because of the relatively small number of subjects in each
group, we restricted our models to only incorporate 10 OTUs. Despite
this restriction, the models performed well (adenoma AUC range = 0.69 -
0.92, advanced adenoma AUC range = 0.80 - 1.00, carcinoma AUC range =
0.82 - 0.98). Interestingly, the 10 OTUs that were used for each model
had little overlap with each other {[}Figure 2{]}. These results support
the earlier community-wide analysis where we observed that the treatment
had an impact on the overall community structure; however, the effect of
treatment was not consistent across patients and diagnosis groups.

@@

Patients diagnosed with adenoma and advanced adenoma, however, did not
experience a shift towards a community structure that resembled those
with normal colons. This may be due to the fundamental differences
between the features of adenomas and advanced adenomas and carcinoma.
Specifically, carcinomas may create an inflammatory milieu that would
impact the structure of the community and removal of that stimulus would
alter said structure. It is possible that the difference between the
microbiota of patients with adenoma and advanced adenoma and those with
normal colons is subtle. This is supported by the reduced ability of our
models to correctly classify patients with adenomas and advanced
adenomas relative to those diagnosed with carcinomas {[}Figure S1{]}.
Given the irregular distribution of microbiota across patients in the
different diagnosis groups, it is possible that we lacked the
statistical power to adequately characterize the change in the
communities following treatment.

There was a subset of patients (6 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 23.08\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 6 individuals (3 with Stage II and 3 with Stage III) was
similar to the distribution observed among the other 20 patients. We
also hypothesized that we may have sampled these patients later than the
rest and their communities may have reverted to a carcinoma-associated
state; however, there was not a statistically significant difference in
the length of time between sample collection among those whose
probabilities increased (358 (121) days) or decreased (334 (143) days)
(Wilcoxon Test; P-value = 0.56) (all days data displayed as median
(IQR)). Finally, it is possible that these patients may not have
responded to treatment as well as the other 20 patients diagnosed with
carcinoma and so the microbiota may not have been impacted the same way.
Again, further studies looking at the role of the microbiota in
recurrence are needed to understand the dynamics following treatment.

Our final hypothesis was that the specific type of treatment altered the
structure of the microbiome. The treatment to remove adenomas and
advanced adenomas was either polyp removal or surgical resection whereas
it was surgical resection alone or in combination with chemotherapy or
with chemotherapy and radiation for individuals with carcinoma. Because
chemotherapy and radiation target rapidly growing cells, these
treatments would be more likely to cause a turnover of the colonic
epithelium driving a more significant change in the structure of the
microbiota. Although, we were able to test for an effect across these
specific types of treatment, the number of patients in each treatment
group was relatively small. Finally, those undergoing surgery would have
received antibiotics and this may be a potential confounder. However,
our pre-treatment stool samples were obtained before the surgery and the
post-treatment samples were obtained long after any effects due to
antibiotic administration on the microbiome would be expected to occur
(344 (142) days).

@@

\textbf{\emph{Model Building.}} {[}Comment: This entire section is
pretty confusing. In the first paragraph it jumbles the treatment models
and the diagnosis models{]} The Random Forest {[}33{]} algorithm was
used to create the three models used. The adenoma model classified
normal versus adenoma, the advanced adenoma model classified normal
versus advanced adenoma, and the carcinoma model classified normal
versus carcinoma. The total number of individuals in the treatment
models was 67 individuals. There were a total of 22 individuals in the
treatment adenoma model, 19 individuals in the treatment advanced
adenoma model, and 26 individuals in the treatment carcinoma model
{[}Table 1{]}. The total number of individuals in the normal versus
diagnosis models was 423 individuals {[}Table 2{]}. There were a total
of 239 individuals in the adenoma model, 262 individuals in the advanced
adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
Each model was then applied to our 67-person cohort {[}Table 1{]} which
assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
(advanced adenoma n = 0 and disease free n = 19), pre-treatment
carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
carcinoma (carcinoma n = 1 and disease free n = 25).

The model included only OTU data obtained from 16S rRNA sequencing and
was constructed using the R package caret (v6.0.73). Optimization of the
mtry hyper-parameter involved making 100 different 80/20 (train/test)
splits of the data where normal and adenoma, normal and advanced
adenoma, or normal and carcinoma were represented in the same proportion
within both the whole data set and the 80/20 split. For each of the
different splits, 20 repeated 10-fold cross validation was performed on
the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. All three models had an optimized mtry of 2. The
hyper-parameter, mtry, defines the number of variables to investigate at
each split before a new division of the data was created with the Random
Forest model.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test. Where
multiple comparison testing was appropriate, a Benjamini-Hochberg (BH)
correction was applied {[}35{]} and a corrected P-value of less than
0.05 was considered significant. The P-values reported are those that
were BH corrected. Model rank importance was determined by obtaining the
median MDA from the 100, 20 repeated 10-fold cross validation and then
ranking from largest to smallest MDA.

@@

\textbf{Figure 1: General differences between adenoma, advanced adenoma,
and carcinoma groups after treatment.} A) Thetayc distance from pre
versus post sample within each individual. A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
5.36e-05). Solid black points represent the median value for each
diagnosis group. B) NMDS of the pre and post-treatment samples for the
adenoma group. C) NMDS of the pre and post-treatment samples for the
advanced adenoma group. D) NMDS of the pre and post-treatment samples
for the carcinoma group.

\textbf{Figure 2: The 10 OTUs used to classify treatment for adenoma,
advanced adenoma, and carcinoma.} A) Adenoma OTUs. B) Advanced Adenoma
OTUs. C) Carcinoma OTUs.

\textbf{Figure 3: OTUs common to those models used to differentiate
between patients with normal colons and those with adenoma, advanced
adenoma, and carcinoma.} A) Venn diagram showing the OTU overlap between
each model. B) For each common OTU the lowest taxonomic identification
and importance rank for each model run is shown.

\textbf{Figure 4: Treatment response based on models built for adenoma,
advanced adenoma, or carcinoma.} A) Positive probability change from
initial to follow up sample in those with adenoma. B) Positive
probability change from initial to follow up sample in those with
advanced adenoma. C) Positive probability change from initial to follow
up sample in those with carcinoma.

@@

\textbf{Table 1: Demographic data of patients in the pre and
post-treatment cohort}

@@

\textbf{Table 2: Demographic data of training cohort}

@@

\textbf{Figure S1: ROC curves of the adenoma, advanced adenoma, and
carcinoma models.} A) Adenoma ROC curve: The light green shaded areas
represent the range of values of a 100 different 80/20 splits of the
test set data and the dark green line represents the model using 100\%
of the data set and what was used for subsequent classification. B)
Advanced Adenoma ROC curve: The light yellow shaded areas represent the
range of values of a 100 different 80/20 splits of the test set data and
the dark yellow line represents the model using 100\% of the data set
and what was used for subsequent classification. C) Carcinoma ROC curve:
The light red shaded areas represent the range of values of a 100
different 80/20 splits of the test set data and the dark red line
represents the model using 100\% of the data set and what was used for
subsequent classification.

\textbf{Figure S2: Summary of important OTUs for the adenoma, advanced
adenoma, and carcinoma models.} A) MDA of the most important variables
in the adenoma model. The dark green point represents the mean and the
lighter green points are the value of each of the 100 different runs. B)
Summary of Important Variables in the advanced adenoma model. MDA of the
most important variables in the SRN model. The dark yellow point
represents the mean and the lighter yellow points are the value of each
of the 100 different runs. C) MDA of the most important variables in the
carcinoma model. The dark red point represents the mean and the lighter
red points are the value of each of the 100 different runs.

\textbf{Figure S3: Pre and post-treatment relative abundance of CRC
associated OTUs within the carcinoma model.}
